Isomorphisms as Groupoids: \\ Homotopy Type Theory with Fractional Types

### 2013-10-13 f.tex
small updates to f.tex

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{color, colortbl}
\usepackage{courier}

@@

@@

@@

@@

@@

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

@@

Because we are \emph{not} in the context of the full $\lambda$-calculus,
which allows arbitrary duplication and erasure of information, values of
fractional types are first-class values that can flow anywhere. The
information-preserving computational infrastructure guarantees that, in a
complete program, every constraint imposed by a fractional value will also be
satisfied exactly once. This property is shared with systems that are based
on linear logic.

@@

Consider a similar algebraic manipulation involving fractional types.
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire2.pdf}
}
\end{center}

@@

One understanding of quantum computing is that it exploits the laws of physics
to build faster machines (perhaps). Another more foundational understanding
is that it provides a computational interpretation of physics, and in
particular directly addresses the question of interpretation of quantum
mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
uses continuations to implement the transactional interpretation of quantum
mechanics~\cite{transactional} which includes as its main ingredient a
fixpoint calculation between waves or particles traveling forwards and
backwards in time. Our work sheds no light on whether this interpretation is
the ``right one'' but it is interesting that we can directly realize it using
the primitives of our language.

The multiplicative structure has a direct connection to entangled quantum
particles, or perhaps entangled particles and anti-particles.  The idea of
entanglement, that an action on one particle is ``instantaneously''
communicated to the other, is analogous to how unifying one value affects its
dual pair which is possibly in another part of the computation.  Again our
model sheds no light on whether this is related to how nature computes but it
is again interesting that we can directly realize the idea.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Speculative Connections}

Negative types, square root types, and imaginary types have also appeared in
the literature: we proceed with a simple explanation. We have so far extended
the set of types to be the positive rational numbers. Now we will push this
and extend the set of types to algebraic numbers. In other words, we will
allow datatypes defined by arbitrary polynomials and allow the roots of such
polynomials to be types.

Consider first an example in which we want to compute with the sides of a
rectangle whose area is 91 and whose length is longer than its width by 6
units. One can solve the quadratic equation to determine that the sides are 7
and 13 and proceed. This however prematuraly forces us to globally solve the
constraint. Instead we can let the two sides of the rectangle be $x$ and
$x+6$ and use the following equation to capture the desired constraint:
\[
x^2 + 6x - 91 = 0
\]
The equation introduces an isomorphism between the type $x^2 + 6x - 91$ and
the type $0$. We can now proceed to compute with the unknown $x$, being
assured that in a closed program, our computation will eventually be
consistent with the solution of the algebraic equation.

Previously the most famous example of a similar nature is the puzzling
isomorphism that one can establish between seven binary trees and one.
A binary tree is defined by the datatype
\[
x = 1 + x * x 
\]
which can be rearranged to the polynomial equation $x^2 - x + 1 = 0$. By
algebraic manipulation we can reason as follows:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make sense
type theoretically but states that even though some of the intermediate steps
make no sense, the final equivalence is valid and can be used to actually
construct an isomorphism between $x^7$ the type of seven binary trees 
and $x$ the type of binary trees.

Discussion of possible polynomials:
\begin{itemize}
\item $b=1+1$. Boring.
\item $b=b+1$. That introduces the natural numbers. No solution for this
  polynomial over the algebraic numbers. We must extend the numbers with
  $\omega$ to get a solution. We reject this in this paper and prefer to
  stick to algebraic numbers. The advantage is that all the isomorphisms are
  valid numerically. With the above type we could subtract $b$ from both
  sides to show that $0=1$ which is nonsense. We can however have infinite
  types as long as they have algebraic solutions.
\item $b^2=2$ or $b = \pm \sqrt{2}$. We have introduced the square root of a
  boolean! If we had superpositions, we could then write a function that
  performs the square root of negation in the sense that applying it twice
  would be equivalent to boolean negation.
\end{itemize}

\paragraph*{Geometry of Interaction (GoI).}
Geometry of Interaction was developed by Girard~\cite{girard1989geometry} as
part of the development of linear logic. It was given a computational
interpretation by Abramsky and
Jagadeesan~\cite{Abramsky:1994:NFG:184662.184664}, and was developed into a
reversible model of computing by
Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. Preliminary investigations
suggest that many of the GoI machine constructions can be simulated by
treating Mackie's bi-directional wires as pairs of wires and replacing the
machine's global state with a typed value on the wire that captures the
appropriate state. This connection is exciting because when viewed through a
Curry-Howard lens it suggests that the logical interpretation of would be a
linear-like logic with a notion of resource preservation and with a natural
computational interpretation.

\paragraph*{Computing in the Field of Algebraic Numbers.}
\label{sec:algebraic-field}
Algebraically, the addition of fractional types corresponds to a move from a
ring-like structure to a full field. Our language captures the structure of
one particular field: that of the rational numbers. As we have seen,
computation in this field is quite expressive and interesting and yet, it has
two fundamental limitations. First it cannot express any recursive type, and
second it cannot express any datatype definitions. We believe these to be two
orthogonal extensions: recursive types were considered in our previous
paper~\cite{James:2012:IE:2103656.2103667}; arbitrary datatypes are however
even more exciting that plain rationals as each datatype definition can be
viewed as a polynomial (see below) which essentially means that we start
computing in the field of algebraic numbers, which includes square roots and
imaginary numbers. As crazy as it might seem, the type $\sqrt{2}$ and even
the type $(1/2)+i(\sqrt{3}/2)$ ``make sense.''  In fact the latter type is
the solution to the polynomial $x^2-x+1=0$ which if re-arranged looks like
$x=1+x*x$ and perhaps more familiarly as the datatype of binary trees $\mu
x.(1+x*x)$. These types happen to have been studied extensively following a
paper by Blass~\cite{seventrees} which used the above datatype of trees to
infer an isomorphism between seven binary trees and one!

We have confirmed that we can extend our language with the datatype
declaration for binary trees and build a witness for this isomorphism that
works as expected. However not every isomorphism constructed from algebraic
manipulation is computationally meaningful. To understand the issue in more
detail, consider the following algebraically valid proof of the isomorphism
in question:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
The question is why such an algebraic manipulation makes sense type
theoretically, even though the intermediate step asking for an isomorphism
between $x^6$ and $1$ has no computational context. In our setting, this
isomorphism can be constructed but it diverges on all inputs (in both
ways). This suggests that, in the field of algebraic numbers, some algebraic
manipulations are somehow ``more constructive'' than others.

A related issue is that not all meaningful recursive types are meaningful
polynomials. For instance $\textsf{nat}=\mu x.(1+x)$ implies the polynomial
$x=1+x$ which has no algebraic solutions without appeal to more complex
structures with limits etc.

@@

\paragraph*{Declarative Continuations.} 
In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which call-by-value is dual to call-by-name and values are dual to
continuations. In more detail, the symmetric calculus contains a ``value''
fragment and a ``continuation'' fragment which are mirror images. Pairs and
sums are treated as duals in the sense that the ``value'' fragment includes
pairs whose mirror image in the ``continuation'' fragment are sums. In
contrast, our language includes pairs and sums in the value fragment and two
symmetries: one that maps the pairs to fractions and another that maps the
sums to subtractions.

\paragraph*{The Duality of Computation.}
The duality between call-by-name and call-by-value was further investigated
by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. Curien and
Herbelin~\cite{Curien:2000} also introduce a calculus that exhibits
symmetries between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. This is to be contrasted with our interpretation of a value of that type
as \emph{either} a value of $A$ or a demand for a value of type $B$. This
calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives no
interpretation to the subtraction connective and like the original symmetric
calculus of Filinski, introduces a duality that relates sums to products and
vice-versa.

\paragraph*{Subtractive Logic.} 
Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
logic which contains a dual to implication. Her work has been distilled in
the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.  In more detail,
Crolard explains the type $A-B$ as the type of \emph{coroutines} with a local
environment of type $A$ and a continuation of type $B$. The description is
complicated by what is essentially the desire to enforce linearity
constraints so that coroutines cannot access the local environment of other
coroutines.

\paragraph*{Negation in Classical Linear Logic} 
Filinski~\cite{Filinski92} uses the negative types of linear logic to model
continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
negative types of linear logic as \emph{acceptors}, which are like
continuations in the sense that they take an input and return no
output. Acceptors however are also similar in flavor to logic variables:
they can be created and instantiated later once their context of use is
determined. Although a formal connection is lacking, it is clear that, at an
intuitive level, acceptors are entities that combine elements of our negative
and fractional types.

\paragraph*{The Lambek-Grishin Calculus.} The ``parsing-as-deduction'' style
of linguistic analysis uses the Lambek-Grishin calculus with the following
types: product, left division, right division, sum, right difference, and
left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division and
difference types are similar to our types but because the calculus lacks
commutativity and associativity and only has limited notions of
distributivity, each connective needs a left and right version. The
Lambek-Grishin exhibits two notions of symmetry but they are unrelated to our
notions. In particular, the first notion of symmetry expresses commutativity
and the second relates products to sums and divisions to subtractions. In
contrast, our two symmetries relate sums to subtractions and products to
divisions.
